Understanding Implications of Dataset Choice for Feature Effect Estimation: A Simulation-Based Investigation through Error Decomposition

### 2025-02-07 paper/paper.tex
paper: add related works and rewrite intro

\section{Introduction}
Most Machine Learning (ML) models can be considered black boxes --- opaque
systems that intrinsically do not allow insight into their internal reasoning,
making it often impossible to explain their decisions. However, this can be
problematic in many domains and applications, such as the healthcare, legal, or
finance sectors, where decisions must be transparent and
accountable~\cite{adadi_peeking_2018}.

Interpretability is crucial to enhance
trust~\cite{ribeiro_why_2016,teach_analysis_1981}, address potential
biases~\cite{guidotti_survey_2019}, fairness and ethical concerns
~\cite{lipton_mythos_2018}, and ensure compliance with regulations such as the
EU's General Data Protection Regulation (GDPR)~\cite{gdpr2016} and AI
Act~\cite{euaia2024}. To address these challenges, the field of Explainable AI
/ Interpretable ML has emerged~\cite{adadi_peeking_2018}. Although there are
many different methods\footnote{For an overview of Explainable AI methods, see
    e.g.~\cite{adadi_peeking_2018,kamath_introduction_2021,molnar_interpretable_2022}.},
we will focus on feature effect methods like \textit{Partial Dependence Plots
    (PDP)}~\cite{friedman_greedy_2001} and \textit{Accumulated Local Effects
    (ALE)}~\cite{apley_visualizing_2020}.

Due to the severity of many applications, it is crucial to utilize these
explainability methods correctly. In general, there are many pitfalls to be
aware of~\cite{molnar_general_2022}, including whether to compute explanations
in-sample, i.e.\ on training data, or out-of-sample, which we refer to as
validation data in the following. In loss-based methods such as
\textit{Permutation Feature Importance (PFI)
}\cite{breiman_random_2001,fisher_all_2019}, this choice is crucial and has
already been studied (e.g., in~\cite{molnar_general_2022}). Nevertheless, other
explainability methods, such as the \textit{Mean Decrease in Impurity (MDI)}
(or \textit{Gini Importance}) of Random Forests, or \textit{SHAP}
values~\cite{loecher_debiasing_2022,loecher_debiasing_2024}, have also been
found to exhibit biases when computed on training data.

However, to the best of our knowledge, there exist no similar studies for
feature effect methods like PDP and ALE. Existing works, including the original
papers of PDP and ALE, rely on training data without further justification
(e.g.,~\cite{apley_visualizing_2020,friedman_greedy_2001,molnar_interpretable_2022}).
In contrast, practitioners often advocate for using unseen test/validation
data\footnote{see, e.g.\ \url{https://github.com/SauceCat/PDPbox/issues/68}
    (10/27/2024)}, or base their choice on practical constraints like dataset
size\footnote{see, e.g.
    \url{https://forums.fast.ai/t/partial-dependence-plot/98465} (10/27/2024)}.
While the training set is usually larger and might thus lead to less variance
in the feature effect estimates, a too large dataset can increase computation
times substantially, particularly for the PDP~\cite{friedman_greedy_2001}. On
the other hand, although feature effects are not based on generalization error
like PFI, it is not clear how much they are affected by overfitting or
distribution shifts between training and validation data.

In this paper, we aim to answer this largely unaddressed, fundamental
methodological question of whether to use training or validation data to
compute feature effects. We perform an empirical simulation study, comparing
feature effect error and uncertainty for PDP and ALE across training data,
validation data, and cross-validation scenarios, considering various data
scenarios and model types. Our main contributions in this paper are as follows.
\begin{enumerate}
    \item We shed light on the question of whether to compute feature effects on training
          data, validation data, or in a cross-validated manner, grounded through our
          comprehensive simulation study, considering feature effect error, bias, and
          variance across different models and data scenarios.
    \item We define methods to quantify feature effect error and uncertainty in a
          theoretical framework, building upon previous work in this area.
    \item We provide an overview of commonly used test functions for simulation studies
          in Interpretable ML, including applications and purposes.
\end{enumerate}

These contributions have several important implications: Our empirically
grounded recommendations enable practitioners and researchers to make informed
decisions about which dataset to select for feature effect computation, helping
to understand potential implications of their choices. Additionally, our
framework for evaluating feature effects and our systematic collection of test
functions provide a foundation for future research in Interpretable ML, with
the latter specifically facilitating test function choice for simulation
studies.

The remainder of this paper is structured as follows. In
\textsc{Section~\ref{sec:related-works}}, we introduce the considered feature
effect methods PDP and ALE, as well as related works on feature effect error
and uncertainty, and give an overview of common test functions for simulation
studies. In \textsc{Section~\ref{sec:quantify-fe-u}}, we extend existing works
and give our definitions of feature effect errors and uncertainty. We then
describe the methodology and set-up of our simulation studies in
\textsc{Section~\ref{sec:methodology-set-up}}, present the results in
\textsc{Section~\ref{sec:results}}, and discuss their implications and
limitations in \textsc{Section~\ref{sec:discussion}}. In
\textsc{Section~\ref{sec:conclusion}}, we briefly conclude our work.

\section{Background \& Related Work}\label{sec:related-works}

@@

\noindent Here, $x_C^{(i)}$ are the actual complement feature values from the dataset of $n$ instances.
To plot this function, a grid of $G$ grid points
$\{(x_S^{(g)}, \widehat{PD}_{\hat f, S}(x_S^{(g)}))\}_{g=1}^G$ can be used~\cite{molnar_relating_2023}.
Molnar et al.~\cite{molnar_general_2022} recommend using quantile-based over equidistant grids.

The PDP assumes that the features in $S$ are independent of the features in
$C$. If this is violated, the perturbations may produce unrealistic data points
outside the underlying joint distribution of the data. This extrapolation issue
can cause misleading
interpretations~\cite{molnar_interpretable_2022,molnar_general_2022}.\\

@@

\noindent Other feature effect methods include the \textit{M-Plot (Marginal Plot)},
which, however, suffers from the omitted variable
bias~\cite{apley_visualizing_2020,friedman_greedy_2001},
or \textit{functional ANOVA (fANOVA)}, which decomposes feature effects
into main and interaction effects~\cite{hooker_discovering_2004}.

Furthermore, methods exist that extend previous ones, such as \textit{Robust
    and Heterogeneity-aware ALE (RHALE)}~\cite{gkolemis_rhale_2023}, or
\textit{Accumulated Total Derivative Effect (ATDEV)} plots, which can be
decomposed into ALE and \textit{Accumulated Cross Effects
    (ACE)}~\cite{liu_model_2018}.

In addition to these global feature effect methods, there are regional effect
plots such as \textit{REPID}~\cite{herbinger_repid_2022}, as well as local
methods, including \textit{ICE (Individual Conditional Expectation)}
curves~\cite{goldstein_peeking_2015} or SHAP dependence plots
(e.g.,~\cite{molnar_interpretable_2022}) based on SHAP
values~\cite{lundberg_unified_2017}.

For the remainder of this paper, we will focus on PDP and ALE as the most
popular global feature effect methods and refer to them when speaking of
feature effects.

@@

\noindent Note that these are similar to the approach in~\cite{molnar_relating_2023},
but we do not specify which data points to use for MC integration. The
variance captures both the variance in the model fits and the variance due to
MC integration. To estimate the MC variance, we propose the following estimators:
\begin{equation}
    \widehat{\text{Variance}}_{MC}(x_S; PDP_f, \widehat{PDP}_f) = \frac{1}{K}\sum_{k=1}^K(PDP_f(x_S) - \widehat{PDP}_f^{(k)}(x_S))^2
\end{equation}

\noindent and
\begin{equation}
    \begin{split}
        \widehat{\text{Variance}}_{MC}(x_S; \widehat{PDP}_{\hat{f}}) = \\ \frac{1}{M(K-1)}\sum_{m=1}^M \sum_{k=1}^K
        \left(\widehat{PDP}_{\hat f^{(m)},S}^{(k)}(x_S) - \frac{1}{K}\sum_{k=1}^K \widehat{PDP}_{\hat f^{(m)},S}^{(k)}(x_S)\right)^2
    \end{split}
\end{equation}

@@

\section{Test Functions for Simulation Studies}\label{sec:test-functions}

Test functions play a crucial role in research, e.g.\ for evaluating and
comparing different methodological approaches, or when conducting simulation
studies. This section synthesizes commonly used test functions across various
domains, providing structured guidance for researchers --- particularly those
in Interpretable ML --- in selecting appropriate test functions for their
simulation studies. By examining test functions from different fields and their
purposes of application, we aim to facilitate experimental design decisions.
While this overview is not exhaustive, it highlights key test functions and
their characteristics to help researchers make informed choices for their
specific research needs.

\paragraph{Test Functions in Optimization.}
The field of optimization, where test functions are essential to enable the
assessment and comparison of optimization algorithms, has established a rich
foundation of test functions. A fundamental approach involves using simple
mathematical expressions like the sphere function~\cite{more_testing_1981}.
These basic functions are often combined with more complex ones like Branin or
Rosenbrock to create comprehensive test suites that incorporate important
properties such as nonlinearity, non-separability, and
scalability~\cite{whitley_building_1998}. A notable framework in this domain is
the Comparing Continuous Optimizer (COCO) platform with its Black Box
Optimization Benchmark (BBOB), offering a structured approach to testing
continuous optimization algorithms through artificial test
functions~\cite{hansen_coco_2016}. These classical test function suites are
well-established in optimization and may also serve as a basis for
Interpretable ML researchers. However, it is important to note that the ability
of these artifical test functions to represent complex real-world behavior is
limited~\cite{zaefferer_simulation-based_2017}.

Physics-derived functions offer a compelling source of real-world test cases,
with the Feynman Symbolic Regression Database (FSReD) being a prominent
example. FSReD comprises 100 physics equations from the seminal
\textit{Feynman's Lectures on Physics (34--36)}, supplemented by 20 more
challenging equations from other seminal physics texts~\cite{udrescu_ai_2020}.
These equations span diverse physics domains including classical mechanics,
electromagnetism, and quantum mechanics. They involve between one and nine
independent variables and incorporate various elementary functions such as
arithmetic operations, trigonometric functions, and exponentials. From that,
tabular datasets were generated through random sampling from defined value
ranges.

Building upon this foundation,
Matsubara~et~al.~\cite{matsubara_rethinking_2024} addressed several limitations
of the original FSReD. Their enhanced framework introduces a three-tiered
categorization of problems (easy, medium, hard) based on their complexity,
incorporates dummy variables to simulate irrelevant features, and implements
more realistic sampling ranges and strategies. Detailed specifications for all
formulas, including their sampling parameters, are available in their work.

@@

\paragraph{Feature Effect Estimation.}
For the trained models, we estimate the feature effects 
$\widehat{PDP}_{\hat f,S}$ and $\widehat{ALE}_{\hat f,S}$ per feature using
equations~(\ref{eq:pdp-estimate}) and (\ref{eq:ale-estimate-uncentered}). For
Monte Carlo integration, we use for each the full training and validation set
as well as a cross-validation strategy. In the latter case, we use a 5-fold
cross-validation on the combined samples of training and validation set, in
each fold fit the model on four folds, and estimate the feature effects on the
remaining fold. We then average the feature effects over all folds. We also
estimate the ground truth feature effects $\widehat{PDP}_{f,S}$ and
$\widehat{ALE}_{f,S}$ since the complexity of the Feynman equation makes
analytical computation infeasible. For the Monte Carlo integration, we use
10000 samples additionally sampled from the DGP, which introduces only a
negligible additional variance term as we show in (\dots) We use 100 grid
points, defined by the quantiles of the theoretical distribution of $X_S$ to
enable a comparison across feature effects. Additionally, we center the curves
after estimation to have a mean effect of $0$ w.r.t.\ the grid points, and we
omit the first and the last grid point after estimation to avoid boundary
effects especially occuring in ALE plots.

\paragraph{Feature Effect Errors.} To quantify the error of the estimated feature effects, we compute the MSE of
the estimated model feature effects with respect to the estimated ground truth
features effects, as well as their bias and variance. For that, we use the
estimators defined in equations~(\ref{eq:mse-estimator}) to
(\ref{eq:variance-estimator}), using $\widehat{PDP}_{f,S}$ instead of
$PDP_{f,S}$ as ground truth (analogously for ALE). We repeat each
dataset-size-model combination $M=30$ times to estimate the error terms, where
each repetition involves drawing a new training and validation set from the
DGP, refitting the models, and estimating the feature effects. To aggregate the
errors into a single measure per feature effect, we average the errors over the
marginal distribution of $X_S$ by averaging over the grid points.

\subsection{Ablation Experiments}\label{sec:experiment-ablation}
\dots

@@

The optimal hyperparameters (OT) were chosen by tuning the models on a separate
data sample. Training and validation data were sampled independently from the
correspondings DGPs, with a training size of $n_{train}$ and a validation size
of 10000 to get a reliable performance estimate. While this scenario is
unrealistic in practice, it allows as to consider hyperparameters as
``pre-selected'' and avoid costly nested resampling strategies for the
simulation study. Each model was tuned for 200 trials using a
\textit{Tree-structured Parzen Estimator (TPE)}~\cite{bergstra_algorithms_2011}
using validation MSE as objective to minimize (equivalent to maximizing
R2-score).

>>>>

\section{Introduction}
Most Machine Learning (ML) models can be considered black boxes --- opaque
systems that intrinsically do not allow insight into their internal reasoning,
making it often impossible to explain their decisions. However, this can be
problematic in many critical domains and applications, such as the healthcare,
legal, or finance sectors, where decisions must be transparent and
accountable~\cite{adadi_peeking_2018}.

Interpretability in ML is crucial to enhance
trust~\cite{ribeiro_why_2016,teach_analysis_1981}, address potential
biases~\cite{guidotti_survey_2019}, fairness and ethical concerns
~\cite{lipton_mythos_2018}, and ensure compliance with regulations such as the
EU's General Data Protection Regulation (GDPR)~\cite{gdpr2016} and AI
Act~\cite{euaia2024}. To address these challenges, the field of Explainable AI
/ Interpretable ML has emerged~\cite{adadi_peeking_2018}. In the last years, a
wide variety of methods has been developed, including model-specific and
model-agnostic approaches ranging from local feature attributions to global
feature importances and effects\footnote{For an overview of Explainable AI
    methods, see
    e.g.~\cite{adadi_peeking_2018,kamath_introduction_2021,molnar_interpretable_2022}.}.

Due to the severity of many applications, it is crucial to utilize these
explainability methods correctly to avoid misleading or incorrect conclusions.
In general, there are many pitfalls to be aware of~\cite{molnar_general_2022},
including whether to compute explanations in-sample, i.e.\ on training data, or
out-of-sample, which we refer to as validation data in the following. Existing
works have studied the implications of this choice for many methods, including
studies on \textit{Permutation Feature Importance (PFI) } (e.g.,
in~\cite{molnar_general_2022}), \textit{Mean Decrease in Impurity (MDI)} of
Random Forests~\cite{loecher_debiasing_2022}, or \textit{SHAP}
values~\cite{loecher_debiasing_2024}.

However, to the best of our knowledge, there exist no such systematic studies
for feature effect methods like \textit{Partial Dependence Plots
    (PDP)}~\cite{friedman_greedy_2001} and \textit{Accumulated Local Effects
    (ALE)}~\cite{apley_visualizing_2020}. Current literature
(e.g.,~\cite{apley_visualizing_2020,friedman_greedy_2001,molnar_interpretable_2022})
predominantly uses training data without explicit justification, while
practitioners also advocate for using unseen validation data (for details, see
\textsc{Section~\ref{sec:related-works}}). Factors that may influence this
choice include potential biases arising from overfitting, the dataset size, and
computational constraints. Although PDP and ALE are not based on generalization
error like PFI, it is not studied if and how they are affected by overfitting.
In addition, larger datasets may improve the feature effect estimates, but also
increase computation times.\\

\noindent In this paper, we aim to answer the largely unaddressed, fundamental
methodological question of whether to use training or validation data to
compute feature effects. We perform an empirical simulation study, estimating
PDP and ALE on training data, validation data, and in a cross-validated manner
for different models and dataset. We decompose the error of PDP and ALE and
compare the error components to understand the implications of the data choice.
Our main contributions can be summarized as follows:

\begin{enumerate}
    \item We shed light on the question of whether to compute feature effects on training
          data, validation data, or in a cross-validated manner, grounded through our
          comprehensive simulation study, considering feature effect error, bias, and
          variance across different models and datasets.
    \item We extend previous work to provide a theoretical framework for decomposing
          feature effect errors and estimating the corresponding components.
    \item We provide an overview of commonly used test functions for simulation studies,
          including applications and purposes, as guidance for researchers in
          Interpretable ML.
\end{enumerate}

\noindent These contributions have several important implications: Firstly, our empirically
grounded recommendations enable practitioners and researchers to make informed
decisions about which data to choose for feature effect computation, helping to
understand potential implications of their choices. Additionally, our framework
for evaluating feature effects and systematic collection of test functions
provide a foundation for future research in Interpretable ML, with the latter
specifically facilitating test function choice for simulation studies.\\

\noindent The remainder of this paper is structured as follows. In
\textsc{Section~\ref{sec:related-works}}, we provide an overview of related
works on feature effects and test functions for simulation studies. In
\textsc{Section~\ref{sec:background}}, introduce the considered feature
effect methods PDP and ALE, and demonstrate how to decompose the error of these
feature effects, providing corresponding definitions and estimators.
We then describe the methodology and set-up of our simulation studies in
\textsc{Section~\ref{sec:methodology-set-up}}, present the results in
\textsc{Section~\ref{sec:results}}, and conclude our work with a discussion 
of their implications and limitations in \textsc{Section~\ref{sec:conclusion}}.

\section{Related Works}\label{sec:related-works}

\paragraph{Feature Effect Methods.}
Current literature offers various methods for analyzing feature effects in ML.
One of the most popular methods is the \textit{Partial Dependence Plot
    (PDP)}~\cite{friedman_greedy_2001}, which describes the marginal effect of one
or two features on the model prediction. The PDP assumes that the features of
interest are independent of the remaining features. When this assumption is
violated, the method may produce unrealistic data points outside the underlying
joint distribution of the data. This extrapolation issue can cause misleading
interpretations~\cite{molnar_interpretable_2022,molnar_general_2022}.

\textit{M-Plots (Marginal Plots)} try to address this issue by considering the
conditional distribution of the remaining features given the feature of interest
but instead suffer from the omitted variable
bias~\cite{apley_visualizing_2020,friedman_greedy_2001}.

The \textit{Accumulated Local Effects (ALE)} plot~\cite{apley_visualizing_2020}
addresses both issues by accumulating local differences in model predictions
along the feature of interest, computing conditional expectations within small
intervals rather than across the entire feature range.

Alternative approaches include \textit{functional ANOVA (fANOVA)}, which
decomposes feature effects into main and interaction
effects~\cite{hooker_discovering_2004}. Recent extensions have further
addressed limitations of these classical methods. For example, \textit{Robust
    and Heterogeneity-aware ALE (RHALE)}~\cite{gkolemis_rhale_2023} adds
heterogeneity quantification to ALE and improving robustness, or
\textit{Accumulated Total Derivative Effect (ATDEV)} plots can be decomposed
into ALE and \textit{Accumulated Cross Effects (ACE)}~\cite{liu_model_2018}.

Beyond these global feature effect methods, there are regional effect plots
such as \textit{REPID}~\cite{herbinger_repid_2022}, and local methods like
\textit{ICE (Individual Conditional Expectation)}
curves~\cite{goldstein_peeking_2015} or SHAP dependence
plots~\cite{lundberg_local_2020}.

In this paper, we focus on PDP and ALE and refer to them when speaking of
feature effects.

\paragraph{Feature Effect Error and Uncertainty.}
Multiple works have approached measuring uncertainty in feature effects. For
probabilistic ML models, Moosbauer et al.\cite{moosbauer_explaining_2021}
derived model-specific confidence bands for PDPs, while applied studies have
proposed bootstrap-based
approaches\cite{esselman_landscape_2015,grange_using_2019}. While the former
are not model-agnostic, the latter often capture only the variance introduced
by Monte Carlo approximation. Nonetheless, the model variance is another source
of uncertainty in PDPs and ALEs, and accounting for it requires multiple model
fits~\cite{apley_visualizing_2020,molnar_general_2022}.

Molnar et al.~\cite{molnar_relating_2023} give formalizations of PDP (and PFI)
as statistical estimators of ground truth estimands, demonstrate the
decomposition of its mean squared error (MSE) into model bias, model variance,
and Monte Carlo variance, and provide estimators for both Monte Carlo and
overall variance. Our work builds upon and extends this idea.

\paragraph{Data Choice for Feature Effect Estimation.}
For many interpretability methods, the choice whether to compute explanations
in-sample (on training data) or out-of-sample (on validation data) can
significantly impact interpretations: For loss-based methods such as
\textit{Permutation Feature Importance (PFI)
}\cite{breiman_random_2001,fisher_all_2019}, this choice is crucial and has
been extensively studied (e.g., in~\cite{molnar_general_2022}). Similar
concerns have been identified for other explainability methods: both the
\textit{Mean Decrease in Impurity (MDI)} of Random Forests and \textit{SHAP}
values have been shown to potentially exhibit biases when computed on training
data~\cite{loecher_debiasing_2022,loecher_debiasing_2024}.

For feature effect methods, however, the implications of this choice remain
largely unexplored. The original works introducing
PDP~\cite{friedman_greedy_2001} and ALE~\cite{apley_visualizing_2020}, as well
as general introductory literature on Interpretable
ML~\cite{molnar_interpretable_2022}, predominantly use training data without
explicit justification. In contrast, practitioners often advocate for using
holdout data or base their choice on practical constraints such as dataset
size\footnote{for examples, see
    \url{https://github.com/SauceCat/PDPbox/issues/68} and
    \url{https://forums.fast.ai/t/partial-dependence-plot/98465} (both accessed
    10/27/2024)}. A too large dataset can increase computation times substantially,
particularly for PDPs~\cite{friedman_greedy_2001}. Moreover, Molnar et
al.~\cite{molnar_relating_2023} estimate the PDP on holdout data when aiming to
quantify the variance of feature effect estimates. Nonetheless, a systematic
study on the implications of the data choice for feature effect estimation is
missing so far.

\subsection{Test Functions for Simulation Studies}\label{sec:test-functions}

Test functions play a crucial role in research, e.g.\ for evaluating different
methodological approaches, or when conducting simulation studies. In the
following, we synthesize commonly used test functions across various domains,
providing structured guidance for researchers --- particularly those in
Interpretable ML --- in selecting appropriate test functions for their
simulation studies. By examining test functions from different fields and their
purposes of application, we aim to facilitate experimental design decisions
rather than giving an exhaustive overview.

\paragraph{Test Functions in Optimization.}
The field of optimization, where test functions are essential to enable the
assessment and comparison of optimization algorithms, has established a rich
foundation of test functions. A fundamental approach involves using simple
mathematical expressions like the sphere function~\cite{more_testing_1981}.
These basic functions are often combined with more complex ones like Branin or
Rosenbrock to create comprehensive test suites that incorporate important
properties such as nonlinearity, non-separability, and
scalability~\cite{whitley_building_1998}. A notable framework in this domain is
the Comparing Continuous Optimizer (COCO) platform with its Black Box
Optimization Benchmark (BBOB), offering a structured approach to testing
continuous optimization algorithms through artificial test
functions~\cite{hansen_coco_2016}. These classical test function suites are
well-established in optimization and may also serve as a basis for
Interpretable ML researchers. However, the ability of these artifical test
functions to represent complex real-world behavior is often
limited~\cite{zaefferer_simulation-based_2017}.

Physics-derived functions offer a compelling source of real-world test cases,
with the \textit{Feynman Symbolic Regression Database
    (FSReD)}~\cite{udrescu_ai_2020} being a prominent example. FSReD comprises 100
physics equations from the seminal \textit{Feynman's Lectures on Physics
    (34--36)}, supplemented by 20 more challenging equations from other seminal
physics texts. These equations span diverse physics domains and involve a
varying number of variables and various elementary functions such as arithmetic
operations, trigonometric functions, and exponentials. Tabular datasets are
generated through random sampling from defined value ranges.

Matsubara~et~al.~\cite{matsubara_rethinking_2024} addressed several limitations
of the original FSReD. They introduce a three-tiered categorization of problems
(easy, medium, hard) based on their complexity, incorporate dummy variables to
simulate irrelevant features, and implement more realistic sampling ranges and
strategies. Detailed specifications for all formulas, including their sampling
parameters, are available in their work.

\section{Background}\label{sec:background}

@@

\noindent Here, $x_C^{(i)}$ are the actual complement feature values from the dataset of $n$ instances.
To plot this function, a grid of $G$ grid points
$\{(x_S^{(g)}, \widehat{PD}_{\hat f, S}(x_S^{(g)}))\}_{g=1}^G$ can be used~\cite{molnar_relating_2023}.
Molnar et al.~\cite{molnar_general_2022} recommend using quantile-based over equidistant grids.\\

@@

@@

\noindent Note that these are similar to the approach in~\cite{molnar_relating_2023},
but we do not specify which data points to use for MC integration. The
variance captures both the variance in the model fits and the variance due to
MC integration. To estimate the MC variance, we propose the following estimators:
\begin{equation}
    \widehat{\text{Variance}}_{MC}(x_S; PDP_f, \widehat{PDP}_f) = \frac{1}{K}\sum_{k=1}^K(PDP_f(x_S) - \widehat{PDP}_f^{(k)}(x_S))^2
    \label{eq:mc-variance-estimator-groundtruth}
\end{equation}

\noindent and
\begin{equation}
    \begin{split}
        \widehat{\text{Variance}}_{MC}(x_S; \widehat{PDP}_{\hat{f}}) = \\ \frac{1}{M(K-1)}\sum_{m=1}^M \sum_{k=1}^K
        \left(\widehat{PDP}_{\hat f^{(m)},S}^{(k)}(x_S) - \frac{1}{K}\sum_{k=1}^K \widehat{PDP}_{\hat f^{(m)},S}^{(k)}(x_S)\right)^2
    \end{split}
    \label{eq:mc-variance-estimator-model}
\end{equation}

@@

@@

\paragraph{Feature effect estimation.}
For the trained models, we estimate the feature effects $\widehat{PDP}_{\hat
        f,S}$ and $\widehat{ALE}_{\hat f,S}$ per feature using
equations~(\ref{eq:pdp-estimate}) and (\ref{eq:ale-estimate-uncentered}). For
Monte Carlo integration, we use for each the full training and validation set
as well as a cross-validation strategy. In the latter case, we use a 5-fold
cross-validation on the combined samples of training and validation set, in
each fold fit the model on four folds, and estimate the feature effects on the
remaining fold. We then average the feature effects over all folds. We also
estimate the ground truth feature effects $\widehat{PDP}_{f,S}$ and
$\widehat{ALE}_{f,S}$ since the complexity of the Feynman equation makes
analytical computation infeasible. For the Monte Carlo integration, we use
10000 samples additionally sampled from the DGP, which introduces only a
negligible additional variance term as we show in (\dots) We use 100 grid
points, defined by the quantiles of the theoretical distribution of $X_S$ to
enable a comparison across feature effects. Additionally, we center the curves
after estimation to have a mean effect of $0$ w.r.t.\ the grid points, and we
omit the first and the last grid point after estimation to avoid boundary
effects especially occuring in ALE plots.

\paragraph{Feature effect errors.} To quantify the error of the estimated feature effects, we compute the MSE of
the estimated model feature effects with respect to the estimated ground truth
features effects, as well as their bias and variance. For that, we use the
estimators defined in equations~(\ref{eq:mse-estimator}) to
(\ref{eq:variance-estimator}), using $\widehat{PDP}_{f,S}$ instead of
$PDP_{f,S}$ as ground truth (analogously for ALE). We repeat each
dataset-size-model combination $M=30$ times to estimate the error terms, where
each repetition involves drawing a new training and validation set from the
DGP, refitting the models, and estimating the feature effects. To aggregate the
errors into a single measure per feature effect, we average the errors over the
marginal distribution of $X_S$ by averaging over the grid points.

\subsection{Ablation Experiments}\label{sec:experiment-ablation}

As described earlier, we conduct two ablation studies for more detailed
insights into the Monte Carlo variance.

\paragraph{Variance decomposition study.} The variance of the feature effects estimated with the procedure described in
the main experiment \textsc{Section~\ref{sec:experiment-main}} consists of two
components: the model variance and the variance due to the Monte Carlo
integration. To disentangle these two sources of variance, we aim to
additionally estimate the Monte Carlo variance using the estimator in
equation~(\ref{eq:mc-variance-estimator-model}). This allows us to substract it
from the total variance estimated in the main experiment and obtain the model
variance, giving more nuanced insights into the implications of the data choice
on feature effect estimation. To obtain this estimate, we maintain the set-up
of the main experiment but for each trained model $\hat f^{(m)}$ (plus five
models from CV $\hat f^{(m,1)} - \hat f^{(m,5)}$), we draw $K=30$ new training
and validation sets from the DGP and estimate the feature effects on these sets
plus in a cross-validated manner with the already fitted models. This allows us
to estimate the Monte Carlo variance as in
equation~(\ref{eq:mc-variance-estimator-model}). Since this study is
computationally more expensive, we do this only for XGBoost (OT and OF) with
the same datasets and sizes as in the main experiment.

\paragraph{Impact of dataset size on Monte Carlo variance.} To investigate the impact of the dataset size on the Monte Carlo variance, we
estimate the Monte Carlo variance between the analytical ground truth feature
effects and the estimated ground truth feature effects on different dataset
sizes. As estimator for the Monte Carlo variance, we use
equation~(\ref{eq:mc-variance-estimator-groundtruth}). With this approach, we
entirely omit the model and thus eliminate other error sources such as model
bias and variance (as with a perfect model fit). To estimate the ground truth
feature effects, we use 50 different dataset sizes ranging from $10^1$ to
$10^6$ on a logarithmic scale. Note that we only consider the
SimpleNormalCorrelated and Friedman1 datasets for this study as an analytical
computation of the FeynmanI.29.16 feature effects is infeasible. For other
details, we maintain the set-up of the main experiment.

@@

The optimal hyperparameters (OT) were chosen by tuning the models on a separate
data sample. Training and validation data were sampled independently from the
correspondings DGPs, with a training size of $n_{train}$ and a validation size
of 10000 to get a reliable performance estimate. While this scenario is
unrealistic in practice, it allows as to consider hyperparameters as
``pre-selected'' and avoid costly nested resampling strategies for the
simulation study. Each model was tuned for 200 trials using a
\textit{Tree-structured Parzen Estimator (TPE)}~\cite{bergstra2011algorithms}
using validation MSE as objective to minimize (equivalent to maximizing
R2-score).
